---
title: "DESeq2 Plots"
subtitle: "Trophoblast organoids collected at passage 3"
author: "Data analysis: Paulo Jannig | Karolinska Institutet"
date: "`r paste('Last update:', format(Sys.time(), '%B %d, %Y'))`"
output: 
  html_document:
    toc: true
    toc_float: 
      collapsed: false
    toc_depth: 3
    number_sections: true
    theme: spacelab
    highlight: default
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
## set work dir as this script's dir
#setwd(dirname(rstudioapi::getSourceEditorContext()$path))

```


```{r}
# define variables ==========================================
filename_prefix <- "" # leave empty if not necessary
species <- "human" # mouse, human, pig, rat
ensembl_version <- 102 # define ensembl release version

file_formats <- list(
  #"png",
  "pdf"
)

# load libraries
required_Packages_Install <- c("tidyverse",
                               "pheatmap",
                               "ggtext",
                               "RColorBrewer",
                               "scico",
                               "writexl",
                               "ggrepel",
                               "ggvenn",
                               "scales"
                               )


for(Package in required_Packages_Install){
  if(!require(Package,character.only = TRUE)) { 
    BiocManager::install(Package, dependencies=TRUE)
  }
  library(Package,character.only = TRUE)
}

# source custom functions
source("load_functions.R") # source custom functions
```

# Import data

## Normalized counts
```{r}
normalized_counts <- readRDS("../02_results/deseq2/deseq2_normalized_counts.Rds")
```

## Metadata
```{r}
metadata <- readRDS("../01_metadata/metadata.Rds")
colnames(metadata)
metadata$sample_id
unique(metadata$Group)
```
## Contrasts
```{r}
high_BMI_vs_Control <- readRDS("../02_results/deseq2/deseq2_high_BMI_vs_Control_lfcShrink.RDS")

T1D_vs_Control <- readRDS("../02_results/deseq2/deseq2_T1D_vs_Control_lfcShrink.RDS")

contrasts <- list(
  high_BMI_vs_Control = high_BMI_vs_Control,
  T1D_vs_Control = T1D_vs_Control
)
```

```{r}

high_BMI_vs_Control.sig <- high_BMI_vs_Control %>% filter(padj<0.05)
T1D_vs_Control.sig <- T1D_vs_Control %>% filter(padj<0.05)

```



## Heatmaps

define color palette
```{r}
# my_palette <- c(
#   "" = "#666666",
#   "" = "#1B9E77",
#   "" = "#D95F02",
#   "" = "#7570B3"
# )

unique(metadata$Group)
#unique(metadata$timepoint)
#unique(metadata$subject)

    # "HIHGNO" = "#1B9E77",
    # "LILGAS" = "#D95F02",
    # "HIHGAS" = "#7570B3"

# Define colors
my_palette_heatmap <- list(
  Group = c(
    Control = "#1B9E77",
    high_BMI = "#D95F02",
    T1D = "#7570B3"
  )
)

```

### all DEGs
for DEGs with padj < 0.05
```{r}
# generate heatmap metadata (annotation_col)
colnames(metadata)
heatmap_metadata <- metadata %>%
  #filter(Group == "RE") %>%
  arrange(Group) %>%
  dplyr::select(c(Sample, Group)) %>%
  mutate(Group = factor(Group,
         levels = c("Control",
                    "high_BMI",
                    "T1D"))) %>%
  column_to_rownames("Sample")

# retrieve genes of interest
#select DEGs for all contrasts (padj < 0.05)
heatmap_genes <- c()
for (contrast in contrasts) {
  genes <- contrast %>%
    filter(padj < 0.05) %>%
    pull(symbol)
  heatmap_genes <- union(heatmap_genes, genes)
  remove(genes, contrast)
}

# # edit color palette
# my_palette_heatmap_RE <- my_palette_heatmap
# my_palette_heatmap_RE$subject <- my_palette_heatmap$subject[grepl(
#   "RE", names(my_palette_heatmap$subject))]

# generate heatmap matrix
heatmap_counts <- normalized_counts %>%
  dplyr::filter(hgnc_symbol %in% heatmap_genes)

heatmap_counts <- heatmap_counts %>%
  dplyr::arrange(desc(matrixStats::rowVars(as.matrix(
    heatmap_counts %>%
      dplyr::select(dplyr::matches(rownames(heatmap_metadata)))
  )))) %>%
  column_to_rownames("hgnc_symbol") %>%
  dplyr::select(dplyr::matches(rownames(heatmap_metadata)))
```

Without gene labels
```{r}
heatmap <- pheatmap::pheatmap(heatmap_counts,
  cluster_rows = T,
  cluster_cols = F,
  scale = "row",
  clustering_distance_rows = "euclidean",
  #cutree_rows = 4,
  show_rownames = F, show_colnames = F,
  cellwidth = 3,
  #cellheight = 4,
  treeheight_row = 5,
  treeheight_col = 15,
  fontsize = 6,
  fontsize_row = 4,
  fontsize_col = 5,
  border_color = NA,
  gaps_col = c(5, 9),
  annotation_colors = my_palette_heatmap,
  annotation_col = heatmap_metadata,
  color = colorRampPalette(
    rev(RColorBrewer::brewer.pal(11, "RdBu"))
  )(256)
)
```

Save plot
```{r}
filename <- "heatmap_significant_zcore_clustered"
for (file_format in file_formats) {
  ggsave(
    plot = heatmap,
    paste0(
      "../03_figures/",
      filename_prefix,
      filename, ".",
      file_format
    ),
    units = "mm",
    dpi = 300,
    width = 100,
    height = 60
  )
}
```

With gene labels
```{r}
heatmap <- pheatmap::pheatmap(heatmap_counts,
  cluster_rows = T,
  cluster_cols = F,
  scale = "row",
  clustering_distance_rows = "euclidean",
  #cutree_rows = 4,
  show_rownames = T, show_colnames = F,
  cellwidth = 3,
  cellheight = 4,
  treeheight_row = 5,
  treeheight_col = 15,
  fontsize = 6,
  fontsize_row = 4,
  fontsize_col = 5,
  border_color = NA,
  gaps_col = c(5, 9),
  annotation_colors = my_palette_heatmap,
  annotation_col = heatmap_metadata,
  color = colorRampPalette(
    rev(RColorBrewer::brewer.pal(11, "RdBu"))
  )(256)
)
```

Save plot
```{r}
filename <- "heatmap_significant_zcore_clustered_labels"
for (file_format in file_formats) {
  ggsave(
    plot = heatmap,
    paste0(
      "../03_figures/",
      filename_prefix,
      filename, ".",
      file_format
    ),
    units = "mm",
    dpi = 300,
    width = 100,
    height = 50
  )
}
```

```{r echo=FALSE}
cat("Saved file :", paste0(
  "../03_figures/",
  filename_prefix, filename,
  ".pdf"
))
```

## MA Plots
```{r}
for (i in seq_along(contrasts)) {
group <- names(contrasts)[i] %>% str_extract(".*(?=_vs_)")
group_ref <- names(contrasts)[i] %>% str_extract("(?<=_vs_).*")
data <- contrasts[[i]]

size_sig <- 0.5
size_ns <- 0.2
plot_title <- paste0("**", group, " vs ", group_ref,"**")

# scico(3, palette = "navia")
# "#021326" "#408A7F" "#FCF3D8"

color_ns <- "#878787"
color_up <- "#B2182B"
color_down <- "#2166AC"
#color_low = "#3869A8"
#color_high = "#D3D5DA"
color_low = "#67001F"
color_high = "#4393C3"
#color_low = "#FB9A99"
#color_high = "#E31A1C"



ma <- ggplot(data, aes(
  x = log2(baseMean),
  y = log2FoldChange
)) +
  ggplot2::geom_point(
    data = filter(data, padj > 0.05 | is.na(padj)),
    colour = color_ns,
    size = size_ns
  ) +
  ggplot2::geom_point(
    data = filter(data, padj < 0.05) %>%
      arrange(-padj), aes(colour = padj), size = size_sig
  ) +
  xlab("log<sub>2</sub> mean expression") +
  ylab("log<sub>2</sub> fold-change") +
  geom_hline(yintercept = 0, linetype = "dotted") +
  ggtitle(plot_title) +
  gg_theme() +
  scale_color_gradient(
    low = color_low, high = color_high,
    limits = c(0, 0.05), oob = scales::squish
  ) +
  #geom_hline(yintercept = 1, linetype = "dotted") +
  #geom_hline(yintercept = -1, linetype = "dotted") +
  annotate("text",
    x = max(log2(data$baseMean)), 
    #y = max(data$log2FoldChange),
    y = 4,
    label = paste0(
      #group,": ",
      "Increased: ",
      nrow(filter(data, padj < 0.05 & log2FoldChange > 0))
    ),
    hjust = 1, vjust = 1, size = 5 / ggplot2:::.pt
  ) +
  annotate("text",
    x = max(log2(data$baseMean)),
    #y = min(data$log2FoldChange),
    y = -4,
    label = paste0(
      #group_ref,
      "Decreased: ",
      nrow(filter(data, padj < 0.05 & log2FoldChange < 0))
    ),
    hjust = 1, vjust = 0, size = 5 / ggplot2:::.pt
  ) +
  coord_cartesian(ylim = c(-4, 4), expand = T) +
  theme(legend.position = "right") +
  geom_label_repel(
    data = filter(data, padj < 0.05 & log2FoldChange > 0) %>% dplyr::slice(1:10),
    aes(label = hgnc_symbol),
    size = 5 / ggplot2:::.pt,
    force = 2,
    label.padding = 0.1,
    # face
    direction = "both",
    max.overlaps = 10,
    #nudge_x = -2,
    #nudge_y = 1,
    min.segment.length = 0,
    segment.size = 0.25
  )  +
  geom_label_repel(
    # data = filter(data, padj < 0.05) %>% dplyr::slice(1:15),
    data = filter(data, hgnc_symbol%in%gene2plot),
    aes(label = hgnc_symbol),
    size = 5 / ggplot2:::.pt,
    force = 2,
    label.padding = 0.1,
    # face
    direction = "both",
    max.overlaps = 20,
    # nudge_x = 2,
    nudge_y = 1,
    min.segment.length = 0,
    segment.size = 0.25
  ) 
ma

# save plot
filename <- paste0(group, "_vs_", group_ref)

for (file_format in file_formats) {
  ggsave_fixed(
    paste0(
      "../03_figures/",
      "MA_plot_",
      filename, ".",
      file_format
    ),
    units = "mm",
    plot_width = 30,
    plot_height = 30,
    width = 50,
    height = 50
  )
}
}
```

# R session info
```{r}
utils:::print.sessionInfo(sessionInfo()[-8])
```